---
title: "Hypercubes"
author: "Paul M"
date: "12/3/2020"
output: html_document
---

Here you are asked to consider the following situation:

Consider n-dimensional unit 'cubes':

* n=1 -> line. Coordinates= (x1) [or (x)].
* n=2 -> a square. Coordinates= (x1,x2) [or (x,y)]. 
* n=3 -> a cube. Coordinates= (x1,x2,x3) [or (x,y,z)]. 
* n=4 -> 4-dimensional hypercube. Coordinates= (x1,x2,x3,x4). 
* n=5 -> 5-dimensional hypercube. Coordinates= (x1,x2,x3,x4,x5). 
* n=N -> N-dimensional hypercube. Coordinates= (x1,x2,x3,x4,???,xN).

Let's suppose that all are unit cubes, so $0<i<1$, for all i.

You are asked to address the following questions:

#### Question 1: 
If we sample a point uniformly at random from inside a hypercube, what is the probability that the point is within a distance of 0.1 of the surface?

#### Question 2: 
How can you use your answer to question 1 to find the what proportion of the volume of an n-dimensional hypercube that is within a distance of 0.1 of the surface?

---

Pseudocode for this problem can be found below.


First, set-up some global variables:
```{r}
set.seed(44)  # set the seed for the random number generator to any number you like
NumberOfTrials <- 50000   # This is how many trials we will run
TotalCountNearSurface <- numeric()
MaxNumberOfDimensions <- 20  # we will go up to this many dimensions
```


Set up a loop to run through the different dimensions, and then for each dimension simulate the procxess of picking a random point. As you do that, keep track of how many fall close to the edge.

```{r}
HowManyCloseToSurface <- 0 # A variable to record how often the point is close to the surface
for (j in 1:MaxNumberOfDimensions) {
  HowManyCloseToSurface <- 0   # reset your counter
  for (i in 1:NumberOfTrials) {
    # pick a point at random in the cube
    MyPoint<-runif(j,0,1)
    #cat(MyPoint)
    #cat("     Max/min: ",max(MyPoint),min(MyPoint<0.1),(max(MyPoint)>0.9)||(min(MyPoint<0.1)))
    #cat("\n")

    # check whether it is within 0.1 of the surface (how do you check this?)
   if ( (max(MyPoint)>0.9) || (min(MyPoint)<0.1)){
      # if it is, increase the value of HowManyCloseToSurface by 1
      HowManyCloseToSurface <- HowManyCloseToSurface+1
   }
  }

  # record the value of HowManyCloseToSurface for this dimension of cube, so that we can plot them all later
  TotalCountNearSurface[j] <- HowManyCloseToSurface
  # the proportion of the volume of the cube that is within 0.1 of the surface
  cat("\nEstimatedProb.:", TotalCountNearSurface[j]/NumberOfTrials)
}


# plot your estimates of the proportion of the volume that is within 01 of the surface (y-axis)
# against the number of dimensions the hypercube has (x-axis)
plot(TotalCountNearSurface/NumberOfTrials)



```

Now plot your estimates of the proportion of the volume that is within 0.1 of the surface (on the y axis)
against the number of dimensions the hypercube has (on the x-axis) and answer thew earlier questions.
```{r}
plot(TotalCountNearSurface/NumberOfTrials)
```

What do we think the answer should be?
```{r}
ExpectedAnswer<-numeric()
for (i in 1:MaxNumberOfDimensions){
  ExpectedAnswer[i]<-1-(0.8^i)
}
# super-impose this on our plot
plot(TotalCountNearSurface/NumberOfTrials)
lines(ExpectedAnswer,col="red")
```


##### Extra question. What do you notice about this proportion as N increases? What value do you think it takes for very large N?

It's pretty clear that the proportion is tending to 1 as $N \rightarrow \infty$. The intuition for this is that the point has to avoid being close to the edge in every dimension, and its position is independent in each dimension, so it becomes harder and harder to do this as the number of dimensions increases.
