---
title: "PiUsingCircle"
author: "Paul M"
date: "12/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Pseudocode for estimating pi using the circle method
```{r}
set.seed(2053)   # remember to set the random number seed so that you can reproduce your output if you end up needing to hunt bugs

# Let's assume the square has length 2, and is centered at (0,0)
# Set circle radius
CircleRadius<-1

NumberOfTrials<-50000    # how many random points to generate
NumberOfPointsInsideCircle<-0   # this will count how often the point lands inside the circle
```


Do the experiment - here's one way to do it. Let's time the different ways to see how long they take.
```{r method1}
# Start the clock
ptm <- proc.time()
for (i in 1:NumberOfTrials){
  # generate a random point in the square
  XCoordinate<-runif(1,-1,1) 
  YCoordinate<-runif(1,-1,1)
  
  # check whether it falls within the circle (i.e. the distance to the origin is less than Circle Radius). If so,
  if (XCoordinate^2+YCoordinate^2<CircleRadius^2){
    NumberOfPointsInsideCircle<-NumberOfPointsInsideCircle+1
  }
  
}
ProportionOfPointsInCircle<-NumberOfPointsInsideCircle/NumberOfTrials

# and then do the math to produce you estimate of pi - yours to write
PiEst <- 4*NumberOfPointsInsideCircle/NumberOfTrials
cat("\nOur estimate of Pi is (method 1): ",PiEst)

# How long did it take?
TimeTaken <- proc.time()[3] - ptm[3]
cat("\nTotal time: ",TimeTaken)
```


Here's another way - is it quicker than the other version?
```{r method2}
# Start the clock
ptm <- proc.time()
AllXCoordinate<-runif(NumberOfTrials,-1,1) 
AllYCoordinate<-runif(NumberOfTrials,-1,1)

# Now check what proportion of these points fall within the circle - yours to write
NumberOfPointsInsideCircle <- sum(AllXCoordinate^2+AllYCoordinate^2<CircleRadius^2)

# and then do the math to produce you estimate of pi - yours to write
PiEst <- 4*NumberOfPointsInsideCircle/NumberOfTrials
cat("\nOur estimate of Pi is (method 2): ",PiEst)
  
# Stop the clock
# How long did it take?
TimeTaken <- proc.time()[3] - ptm[3]
cat("\nTotal time: ",TimeTaken)
```

If you are feeling fancy, write some code to plot your points and the circle and square, 
or plot your current estimate of pi as you perform the iterations

Now you could write some code to repeat the above for different sizes of circle (i.e. different values of CircleRadius). This would be easiest if you turned the above into a function. Does it matter what size you use in terms of efficiency of the algorithm?


